---
title: "Reproducible Research -- Peer Assessment 1"
author: 'By: Sakinah'
date: "November 14, 2015"
output: 
  html_document: 
    toc: yes
---

## Reproducible Research -- Peer Assessment 1
### Loading and preprocessing the data
Reading data from excel

```{r}
mydata <- read.csv("activity.csv")
```

###The summary of the data

```{r}
summary(mydata)
```


```{r,echo=FALSE}
CompleteData<-complete.cases(mydata)  
# number of missing data 
nMissingData <- length(CompleteData[CompleteData==FALSE])  
# number of Complete data 
nCompleteData <- length(CompleteData[CompleteData==TRUE])  
title="Boxplot : Missing vs. Complete Cases"  
barplot(table(CompleteData),main=title,xaxt='n' ) 
axis(side=1,at=c(.7,1.9),labels=c("Missing Data","Complete Data"),tick=TRUE)           
# label the barplot  
text(.7,0,labels=nMissingData, pos=3)                                            
text(1.9,0,labels=nCompleteData, pos=3)                      
```

### What is mean total number of steps taken per day?

The **mean** and **median** total number of steps taken
per day:
```{r, echo=FALSE}

CleanData<-subset(mydata,complete.cases(mydata))
GroupByDay<-split(CleanData,CleanData$date,drop=TRUE)
DailySteps<-sapply(GroupByDay,function(x) sum(x$steps))

mean(DailySteps)
median(DailySteps)

# plot a histogram 
hist(DailySteps, main="Histogram Total Steps per Day", xlab="# Steps")  
abline(v=mean(DailySteps), lty=3, col="blue")                    
abline(v=median(DailySteps), lty=4, col="red")                  
text(mean(DailySteps),25,labels="mean", pos=4, col="blue")      #
text(median(DailySteps),23,labels="median", pos=4, col="red")    
summary(DailySteps)
```

### What is the average daily activity pattern?

```{r, echo=FALSE}
SplitByInterval<-split(CleanData,CleanData$interval,drop=TRUE)
intervalAverage<-sapply(SplitByInterval,function(x) mean(x$steps))
plot(intervalAverage, type = "l",
     main = "Interval", 
     ylab = "Average",
     xlab = "Interval")
abline(v=which.max(intervalAverage), lty=3, col="blue") 
text(which.max(intervalAverage),max(intervalAverage),  
     labels=paste("max = ",as.character(round(max(intervalAverage)))), 
     pos=4, col="blue")
names(which.max(intervalAverage))
round(max(intervalAverage))
which.max(intervalAverage)
```

###Imputing missing values

```{r}
mydata$steps[is.na(mydata$steps)] = mean(mydata$steps, na.rm=TRUE)
ImpuData = mydata
```

```{r, echo=FALSE}
splitNewByDay <- split(ImpuData,ImpuData$date, drop=TRUE)       # split the newData by date  
dailyStepsNew <- sapply(splitNewByDay, function(x) sum(x$steps)) # numeric vector w/ daily sum of steps  
hist(dailyStepsNew, main="Histogram : Total Steps per Day (Imputation data)", xlab=" Number of Steps") # plot a histogram  
abline(v=mean(dailyStepsNew), lty=3, col="blue")                            # draw a blue line thru the mean  
abline(v=median(dailyStepsNew), lty=4, col="red")                           # draw a red line thru the median  
text(mean(dailyStepsNew),35,labels="mean", pos=4, col="blue")               # label the mean  
text(mean(dailyStepsNew),33,labels="median", pos=4, col="red")              # label the median  
summary(dailyStepsNew)
```


Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

**By using the average value for imputation, the only difference is in the frequency or the number of observations for the center bar of the new histogram**
```{r, echo=FALSE}
par(mfrow=c(1,2))

### plot the original histogram
hist(DailySteps, main="With Original Data", xlab="Number of Steps", ylim=c(0,35)) # plot a histogram  
abline(v=mean(DailySteps), lty=3, col="blue")                      # draw a blue line thru the mean  
abline(v=median(DailySteps), lty=4, col="red")                     # draw a red line thru the median  
text(mean(DailySteps),25,labels="mean", pos=4, col="blue")         # label the mean  
text(mean(DailySteps),23,labels="median", pos=4, col="red")        # label the median  

### plot the imputed histogram
hist(dailyStepsNew, main="With Imputation Data", xlab="Number of Steps", ylab="") # plot a histogram  
abline(v=mean(dailyStepsNew), lty=3, col="blue")                      # draw a blue line thru the mean  
abline(v=median(dailyStepsNew), lty=4, col="red")                     # draw a red line thru the median  
text(mean(dailyStepsNew),35,labels="mean", pos=4, col="blue")         # label the mean  
text(mean(dailyStepsNew),33,labels="median", pos=4, col="red")        # label the median  
```

###Are there differences in activity patterns between weekdays and weekends?
This person does most of his or her walking on the weekends!
```{r, echo=FALSE}
mydata$date <- as.Date(strptime(mydata$date, format="%Y-%m-%d")) # convert date to a date() class variable  
mydata$day <- weekdays(mydata$date)                              # build a 'day' factor to hold weekday / weekend  
for (i in 1:nrow(mydata)) {                                       # for each day  
  if (mydata[i,]$day %in% c("Saturday","Sunday")) {             # if Saturday or Sunday,
    mydata[i,]$day<-"weekend"                                 #   then 'weekend'
  }
  else{
    mydata[i,]$day<-"weekday"                                 #    else 'weekday'
  }
}

## aggregate newData by steps as a function of interval + day  
stepsByDay <- aggregate(mydata$steps ~ mydata$interval + mydata$day, mydata, mean)

## reset the column names to be pretty & clean
names(stepsByDay) <- c("interval", "day", "steps")

## plot weekday over weekend time series
par(mfrow=c(1,1))  
with(stepsByDay, plot(steps ~ interval, type="n", main="Weekday vs. Weekend Avg."))  
with(stepsByDay[stepsByDay$day=="weekday",], lines(steps ~ interval, type="l", col="chocolate"))  
with(stepsByDay[stepsByDay$day == "weekend",], lines(steps ~ interval, type="l", col="16" ))  
legend("topright", lty=c(1,1), col = c("chocolate", "16"), legend = c("weekday", "weekend"), seg.len=3)

```